Flexible Probabilistic Forecasting with Distributional Trees and Forests

### 2018-04-12 pkg/disttree/inst/MIP-Seminar/slides.Rnw
last touch-ups
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2149 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 900e9ecccfd891c2b1cf9bcba64ae4ac93104716

\begin{frame}[fragile]
\frametitle{Statistical Modeling}
\visible<2,3, 6-> {Assumption:}
\vspace{-0.7cm}
\begin{center}
\begin{tikzpicture}
\visible<2,3, 6-> {\node[] at (0,0.4) {X};}
\visible<4-5>{
\node[inner sep=0pt] (today) at (-1.8,-0.4)
    {\includegraphics[width=.365\textwidth]{today.png}};
\node[font=\scriptsize] at (-1.9,-2) {15.03.2018};
}
\visible<2->{
\draw[->] (0.4,0.4)--(0.9,0.4);
\draw[] (1,0) rectangle (2.5,0.8);
\node[font=\scriptsize] at (1.75,0.4) {nature};
\draw[->] (2.6,0.4)--(3.1,0.4);
}
\visible<2,3, 6->{
\node[] at (3.5,0.4) {Y};
}
\visible<4-5>{
\node[inner sep=0pt] (future) at (5.3,-0.4)
    {\includegraphics[width=.365\textwidth]{future.png}};
\node[font=\scriptsize] at (5.4,-2) {16.03.2018};
}
\end{tikzpicture}
\end{center}
\visible<3, 6->{
\vspace{-2cm}
Approximation of reality based on a set of observations.\\
}
\visible<5>{
\vspace{2cm}
X = observed precipitation of today,\\
\qquad current temperature / humidity / \dots,\\
\qquad physical model output\\
\vspace{0.15cm}
Y = daily precipitation sums\\
}

\vspace{-3.5cm}
\visible<7->{
\textbf{Application/Motivation:}
\begin{itemize}
\item Prediction.
\item Caption of dependencies and interactions.
\end{itemize}
}
\vspace{0.5cm}
\visible<8->{
\textbf{Two main approaches / ``cultures'':}
\begin{itemize}
\item Data modeling.
\item Algorithmic modeling.
\end{itemize}
}
\end{frame}

\begin{frame}[fragile]
\frametitle{Data Modeling}
\vspace{-0.1cm}
\begin{center}
\begin{tikzpicture}
\node[] at (0,0.4) {X};
\draw[->] (0.4,0.4)--(0.9,0.4);
\draw[] (1,0) rectangle (2.5,0.8);
\node[font=\scriptsize] at (1.75,0.4) {f(X; par)};
\draw[->] (2.6,0.4)--(3.1,0.4);
\node[] at (3.5,0.4) {Y};
\end{tikzpicture}
\end{center}
\vspace{-0.4cm}
\only<1-6>{
\begin{itemize}
\item Approximate data generating process up to a few unknown processes.
\item Mathematical equations, model assumptions.
\item Data: estimation of parameters.
\end{itemize}
}
\only<7-10>{
\vspace{1cm}
Summary:
\begin{itemize}
\item<8-10> Stochastic model to approximate data generating process.\\
\only<9-10>{$\Rightarrow$ fixed $f$.\\}
\item<10> Disadvantage: Risk of incorrect assumptions.
\end{itemize}
}
\only<1>{\vspace{1cm}}

@@

\visible<2-6>{
\begin{minipage}{0.65\textwidth}
\textbf{Example:} 
\only<2-3>{Linear model}\only<4>{Generalized additive model}\only<5-6>{Generalized additive model\\ for location, scale, and shape}
\\
\only<2-3>{
$$\visible<3>{\mathbb{E}(Y|X) = } f(X; \alpha) = \alpha_0 + \alpha_1 \cdot X$$  
\vspace{0.15cm}
}
\only<4>{
$$\mathbb{E}(Y|X) = f(X; \alpha) = \alpha + s(X)$$  
\vspace{0.15cm}
}
\only<5-6>{
\vspace{-1.3cm}

\begin{align*}
\visible<6>{\mathbb{E}(Y|X) = }\mu(X) = \alpha_{\mu} + s_{\mu}(X)\\  
\visible<6>{\sqrt{Var(Y|X)} = }\sigma(X) = \alpha_{\sigma} + s_{\sigma}(X)
\end{align*}
}
\end{minipage}
%\vspace{-0.6cm}
%\hfill
\begin{minipage}{0.3\textwidth}
<<motivation_GLM, echo=FALSE, results=hide>>=
nobs <- 200
## GLM
set.seed(7)
x <- c(1:nobs)/nobs
ytrue <- 1+x*1.5
y <- ytrue + rnorm(nobs,0,0.3)
@
\only<2-3>{
<<plot_motivation_GLM_single, fig=TRUE, echo=FALSE>>=
par(mar=c(2,0,2,0))
plot(y=y , x=x, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19)
box(lwd=5)
lines(x = x, y = ytrue, col = pal["forest"], lwd = 7, main = "")
@
}
<<motivation_GAM, echo=FALSE, results=hide>>=
## GAM
set.seed(7)
x <- c(1:nobs)/nobs
x <- 2*(x-0.5)
ytrue <- x^3 
y <- ytrue + rnorm(nobs,0,0.3)
@
\only<4>{
<<plot_motivation_GAM, fig=TRUE, echo=FALSE>>=
par(mar=c(2,0,2,0))
plot(y=y , x=x, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19)
box(lwd=5)
lines(x = x, y = ytrue, col = pal["forest"], lwd = 7, main = "")
@
}
<<motivation_GAMLSS, echo=FALSE, results=hide>>=
## GAMLSS
set.seed(7)
x <- c(1:nobs)/nobs
x <- 2*(x-0.5)
ytrue <- x^3
var <- exp(-(2*x)^2)/2
y <- ytrue + rnorm(nobs, 0, 0.1 + var)
@
\only<5-6>{
<<plot_motivation_GAMLSS, fig=TRUE, echo=FALSE>>=
par(mar=c(2,0,2,0))
plot(x, y, xaxt = "n", yaxt = "n", ann = FALSE, type = "n")
polygon(c(x, rev(x)), c(ytrue + 0.1 + var, rev(ytrue - 0.1 - var)),
  col = pallight["forest"], border = "transparent")
lines(x, ytrue, col = pal["forest"], lwd=7)
points(x, y, col = "slategray", pch = 19)
box(lwd = 5)
@
}
\end{minipage}
}
\end{frame}

@@

\begin{frame}[fragile]
\frametitle{Distributional Tree}
\begin{minipage}{0.78\textwidth}
\begin{center}
\vspace{0.2cm}
\visible<2->{
Model: \code{disttree(y $\sim$ x)}\\
}
\vspace{-0.2cm}
\only<2-3>{
%\setkeys{Gin}{width=0.\linewidth, height=0.68\linewidth}
<<plottree_estpar, fig=TRUE, echo=FALSE, width = 7>>=
# function for output in terminal panels
FUN <- function (x) 
{
  cf <- x$coefficients
  cf <- matrix(cf, ncol = 1, dimnames = list(names(cf), ""))
  c(sprintf("n = %s", x$nobs), "Estimated parameters:", parse(text = paste0("mu == '", format(round(cf[1], 2), nsmall = 2), "'")), 
                                                        parse(text = paste0("sigma == '", format(round(cf[2], 2), nsmall = 2), "'")))
}

@@

## plot version using FUN and tree of class 'disttree'
plot(tree, drop = TRUE, tnex = 1.7, FUN = FUN,
     tp_args = list(fill = paltrees[c(1, 3)], width = 18), 
     ip_args = list(fill = paltrees[c(2, 3)]))
@
}
\only<4->{
%\setkeys{Gin}{width=0.68\linewidth, height=0.68\linewidth}
<<plottree_dens, fig=TRUE, echo=FALSE, width = 7>>=
node_density <- function (tree, xscale = NULL, yscale = NULL, horizontal = FALSE,
                          main = "", xlab = "", ylab = "Density", id = TRUE, rug = TRUE,
                          fill = paltrees[c(1, 3)], col = "black", lwd = 0.5, ...) 
{
  yobs <- tree$data[,as.character(tree$info$formula[[2]])]
  ylines <- 1.5
  if (is.null(xscale)) xscale <- c(-5.1,22.5)
  if (is.null(yscale)) yscale <- c(-0.05,0.45)
  xr <- xscale
  yr <- yscale

@@

plot(tree, tnex = 1.7, drop = TRUE,
     terminal_panel = node_density,
     tp_args = list(fill = paltrees[c(1, 3)]), 
     ip_args = list(fill = paltrees[c(2, 3)]))
@
}
\visible<3->{
R-package \code{disttree}
}
\end{center}
\end{minipage}
\begin{minipage}{0.21\textwidth}

@@

\center
\visible<5->{
Algorithmic \\
Modeling\\
}

@@

\visible<6->{
Data \\
Modeling
}

@@

\visible<7->{
Combined \visible<8->{\checkmark}
}
\end{minipage}
\end{frame}

@@

Schlosser L, Hothorn T, Stauffer R, Zeileis A (2018).
  \dquote{Distributional Regression Forests for Probabilistic Precipitation Forecasting in Complex Terrain.}
  \emph{Manuscript under preparation}.

>>>>

\begin{frame}[fragile]
\frametitle{Statistical Modeling}
\visible<2,3, 5-> {Assumption:}
\vspace{-0.7cm}
\begin{center}
\begin{tikzpicture}
\visible<2,3, 5-> {\node[] at (0,0.4) {X};}
\visible<4>{
\node[inner sep=0pt] (today) at (-1.8,-0.4)
    {\includegraphics[width=.365\textwidth]{today.png}};
\node[font=\scriptsize] at (-1.9,-2) {2018-03-15};
}
\visible<2->{
\draw[->] (0.4,0.4)--(0.9,0.4);
\draw[] (1,0) rectangle (2.5,0.8);
\node[font=\scriptsize] at (1.75,0.4) {nature};
\draw[->] (2.6,0.4)--(3.1,0.4);
}
\visible<2,3, 5->{
\node[] at (3.5,0.4) {Y};
}
\visible<4>{
\node[inner sep=0pt] (future) at (5.3,-0.4)
    {\includegraphics[width=.365\textwidth]{future.png}};
\node[font=\scriptsize] at (5.4,-2) {2018-03-16};
}
\end{tikzpicture}
\end{center}
\visible<3, 5->{
\vspace{-2cm}
Approximation of reality based on a set of observations.\\
}
\visible<4>{
\vspace{2cm}
X = observed precipitation of today,\\
\qquad current temperature / humidity / \dots,\\
\qquad physical model output\\
\vspace{0.15cm}
Y = daily precipitation sums\\
}

\vspace{-3.5cm}
\visible<6->{
\textbf{Application/Motivation:}
\begin{itemize}
\item Prediction.
\item Caption of dependencies and interactions.
\end{itemize}
}
\vspace{0.5cm}
\visible<7->{
\textbf{Two main approaches / ``cultures'':}
\begin{itemize}
\item Data modeling.
\item Algorithmic modeling.
\end{itemize}
}
\end{frame}

\begin{frame}[fragile]
\frametitle{Data Modeling}
\vspace{-0.1cm}
\begin{center}
\begin{tikzpicture}
\node[] at (0,0.4) {X};
\draw[->] (0.4,0.4)--(0.9,0.4);
\draw[] (1,0) rectangle (2.5,0.8);
\node[font=\scriptsize] at (1.75,0.4) {f(X; par)};
\draw[->] (2.6,0.4)--(3.1,0.4);
\node[] at (3.5,0.4) {Y};
\end{tikzpicture}
\end{center}
\vspace{-0.4cm}
\only<1-5>{
\begin{itemize}
\item Approximate data generating process up to a few unknown processes.
\item Mathematical equations, model assumptions.
\item Data: estimation of parameters.
\end{itemize}
}
\only<6->{
\vspace{1cm}
Summary:
\begin{itemize}
\item Stochastic model to approximate data generating process.\\
\only{$\Rightarrow$ fixed $f$.\\}
\item Disadvantage: Risk of incorrect assumptions.
\end{itemize}
}
\only<1>{\vspace{1cm}}

@@

\visible<2-5>{
\begin{minipage}{0.65\textwidth}
\textbf{Example:} 
\only<2-3>{Linear model}\only<4>{Generalized additive model}\only<5-6>{Generalized additive model\\ for location, scale, and shape}
\\
\only<2-3>{
$$\visible<3>{\mathbb{E}(Y|X) = } f(X; \alpha) = \alpha_0 + \alpha_1 \cdot X$$  
\vspace{0.15cm}
}
\only<4>{
$$\mathbb{E}(Y|X) = f(X; \alpha) = \alpha + s(X)$$  
\vspace{0.15cm}
}
\only<5>{
\vspace{-1.29cm}

\begin{align*}
\visible<5>{\mathbb{E}(Y|X) = }\mu(X) = \alpha_{\mu} + s_{\mu}(X)\\  
\visible<5>{\sqrt{Var(Y|X)} = }\sigma(X) = \alpha_{\sigma} + s_{\sigma}(X)
\end{align*}
}
\end{minipage}
%\vspace{-0.6cm}
%\hfill
\begin{minipage}{0.3\textwidth}
<<motivation_GLM, echo=FALSE, results=hide>>=
nobs <- 200
## GLM
set.seed(7)
x <- c(1:nobs)/nobs
ytrue <- 1+x*1.5
y <- ytrue + rnorm(nobs,0,0.3)
@
\only<2-3>{
<<plot_motivation_GLM_single, fig=TRUE, echo=FALSE>>=
par(mar=c(2,0,2,0))
plot(y=y , x=x, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19)
box(lwd=5)
lines(x = x, y = ytrue, col = pal["forest"], lwd = 7, main = "")
@
}
<<motivation_GAM, echo=FALSE, results=hide>>=
## GAM
set.seed(7)
x <- c(1:nobs)/nobs
x <- 2*(x-0.5)
ytrue <- x^3 
y <- ytrue + rnorm(nobs,0,0.3)
@
\only<4>{
<<plot_motivation_GAM, fig=TRUE, echo=FALSE>>=
par(mar=c(2,0,2,0))
plot(y=y , x=x, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19)
box(lwd=5)
lines(x = x, y = ytrue, col = pal["forest"], lwd = 7, main = "")
@
}
<<motivation_GAMLSS, echo=FALSE, results=hide>>=
## GAMLSS
set.seed(7)
x <- c(1:nobs)/nobs
x <- 2*(x-0.5)
ytrue <- x^3
var <- exp(-(2*x)^2)/2
y <- ytrue + rnorm(nobs, 0, 0.1 + var)
@
\only<5>{
<<plot_motivation_GAMLSS, fig=TRUE, echo=FALSE>>=
par(mar=c(2,0,2,0))
plot(x, y, xaxt = "n", yaxt = "n", ann = FALSE, type = "n")
polygon(c(x, rev(x)), c(ytrue + 0.1 + var, rev(ytrue - 0.1 - var)),
  col = pallight["forest"], border = "transparent")
lines(x, ytrue, col = pal["forest"], lwd=7)
points(x, y, col = "slategray", pch = 19)
box(lwd = 5)
@
}
\end{minipage}
}
\end{frame}

@@

\begin{frame}[fragile]
\frametitle{Distributional Tree}
\begin{minipage}{0.78\textwidth}
\begin{center}
\vspace{0.2cm}
\visible<1->{
Model: \code{disttree(y $\sim$ x)}\\
}
\vspace{-0.2cm}
\only<1>{
%\setkeys{Gin}{width=0.\linewidth, height=0.68\linewidth}
<<plottree_estpar, fig=TRUE, echo=FALSE, width = 7>>=
# function for output in terminal panels
FUN <- function (x) 
{
  cf <- x$coefficients
  cf <- matrix(cf, ncol = 1, dimnames = list(names(cf), ""))
  c(sprintf("n = %s", x$nobs), "Estimated parameters:", parse(text = paste0("mu == '", format(round(cf[1], 2), nsmall = 2), "'")), 
                                                        parse(text = paste0("sigma == '", format(round(cf[2], 2), nsmall = 2), "'")))
}

@@

## plot version using FUN and tree of class 'disttree'
plot(tree, drop = TRUE, tnex = 1.7, FUN = FUN,
     tp_args = list(fill = paltrees[c(1, 3)], width = 18), 
     ip_args = list(fill = paltrees[c(2, 3)]))
@
}
\only<2->{
%\setkeys{Gin}{width=0.68\linewidth, height=0.68\linewidth}
<<plottree_dens, fig=TRUE, echo=FALSE, width = 7>>=
node_density <- function (tree, xscale = NULL, yscale = NULL, horizontal = FALSE,
                          main = "", xlab = "", ylab = "Density", id = TRUE, rug = TRUE,
                          fill = paltrees[c(1, 3)], col = "black", lwd = 0.5, ...) 
{
  yobs <- tree$data[,as.character(tree$info$formula[[2]])]
  ylines <- 1.5
  if (is.null(xscale)) xscale <- c(-5.1,22.5)
  if (is.null(yscale)) yscale <- c(-0.05,0.45)
  xr <- xscale
  yr <- yscale

@@

plot(tree, tnex = 1.7, drop = TRUE,
     terminal_panel = node_density,
     tp_args = list(fill = paltrees[c(1, 3)]), 
     ip_args = list(fill = paltrees[c(2, 3)]))
@
}
\visible<1->{
R-package \code{disttree}
}
\end{center}
\end{minipage}
\begin{minipage}{0.21\textwidth}

@@

\center
\visible<3->{
Algorithmic \\
Modeling\\
}

@@

\visible<4->{
Data \\
Modeling
}

@@

\visible<5->{
Combined \visible<6->{\checkmark}
}
\end{minipage}
\end{frame}

@@

Schlosser L, Hothorn T, Stauffer R, Zeileis A (2018).
  \dquote{Distributional Regression Forests for Probabilistic Precipitation Forecasting in Complex Terrain.}
  \emph{arXiv 1804.02921}, arXiv.org E-Print Archive.
  \url{http://arxiv.org/abs/1804.02921}
